---
title: "Advanced Bioinformatics 2026 assessment"
author: "K25117163"
date: "2026-04-14"
output: html_document
---

### **3. R/RStudio Assessments**
Perform a number of tasks using basic R operations.

#### **Task 3.1: Summing Integers**
```{r}
# Evaluates the sum of integers between 5 and 55
sum(5:55)
```

#### **Task 3.2: Custom Functions**
```{r}
sumfun <- function(n) {
  sum(5:n)
}
# Results for n=10, n=20, n=100
sumfun(10)
sumfun(20)
sumfun(100)
```

#### **Task 3.3: Fibonacci Series and Loops**
```{r}
# Create vector to hold 12 numbers
fib <- numeric(12)
# Set the first two numbers to 1
fib[1:2] <- 1
# Use a for loop to calculate the rest 
for (i in 3:12) {
  fib[i] <- fib[i-1] + fib[i-2]
}
# Print on screen
print(fib)
```

#### **Task 3.4: mtcars Boxplot**
```{r}
library(ggplot2)
# Plot mpg as a function of gears
ggplot(mtcars, aes(x = factor(gear), y = mpg, fill = factor(gear))) +
  geom_boxplot() +
  labs(x = "Number of Gears", y = "Miles per Gallon", fill = "Gears") +
  theme_minimal()
```


#### **Task 3.5: Linear Model**
```{r}
# Reminder car dataset measures distance in feet(ft)
# Fit a linear relationship
fit <- lm(dist ~ speed, data = cars)

# Use summary() to extract coefficients, standard errors, and other statistics
summary_fit <- summary(fit)
print(summary_fit)

# Extract specific values for the report
intercept <- coef(summary_fit)[1, 1]
slope <- coef(summary_fit)[2, 1]
se_intercept <- coef(summary_fit)[1, 2]
se_slope <- coef(summary_fit)[2, 2]

# Displaying results clearly
cat("Fitted Intercept:", round(intercept, 3), "(SE:", round(se_intercept, 3), ")\n")
cat("Fitted Slope:", round(slope, 3), "(SE:", round(se_slope, 3), ")\n")

```

#### **Task 3.6: Linear Fit**
```{r}
# Reminder car dataset measures distance in feet(ft)
# Load the ggplot2 library
library(ggplot2)

# Create the plot
ggplot(cars, aes(x = speed, y = dist)) +
  # Add the raw data points
  geom_point(color = "red", size = 2) +
  # Add the linear regression line (method = "lm")
  # se = TRUE adds the shaded confidence interval
  geom_smooth(method = "lm", color = "green", se = TRUE) +
  # Add descriptive labels and a theme
  labs(title = "Linear Fit: Speed vs. Braking Distance",
       subtitle = "Data source: cars dataset",
       x = "Speed (mph)",
       y = "Braking Distance (ft)") +
  theme_bw()
```

#### **Task 3.7: Estimate average reaction time**
```{r}
# Fit the model: Distance = (ReactionTime * Speed) + (k * Speed^2)
#   We remove the intercept (0 +) because at 0 mph, distance is 0 ft.
fit_physics <- lm(dist ~ 0 + speed + I(speed^2), data = cars)

# Extract the coefficient for 'speed'
#   This coefficient is the reaction time in hours.
rt_hours <- coef(fit_physics)["speed"]

# Convert reaction time to seconds.
#   Speed is in mph, but distance is in feet. 
#   1 mph = 1.467 feet per second.
# Therefore, the coefficient is already in 'seconds' if we adjust for the mph-to-fps conversion.
rt_seconds <- rt_hours * (5280 / 3600) 

# Display results
print(summary(fit_physics))
cat("\nEstimated Average Reaction Time:", round(rt_seconds, 3), "seconds\n")

# Visualisation
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point() +
# Overlay the quadratic fit
geom_smooth(method = "lm", formula = y ~ 0 + x + I(x^2), color = "green") +
  labs(title = "Quadratic Model: Reaction Time + Braking Distance",
       subtitle = paste("Estimated Reaction Time:", round(rt_seconds, 3), "s"),
       x = "Speed (mph)",
       y = "Stopping Distance (ft)") +
  theme_minimal()
```


#### **Task 3.8: LMS_RNAseq**
```{r}
## Define filenames
file_counts <- "exercise1_counts.csv"
file_info <- "exercise1_sample_description.info"

# Read the data
counts <- read.csv(file_counts, row.names = 1)
info <- read.table(file_info, header = TRUE)

# Verify import
print("Dimensions of Counts (Genes x Samples):")
print(dim(counts))
print("Dimensions of Info (Samples x Metadata):")
print(dim(info))
```

#### **Task 3.9: Dimensions**
```{r}
# Conversion
col_data <- as.data.frame(info)

# Check the dimensions of both objects
# col_data rows = counts columns
cat("Dimensions of col_data (Rows x Cols):", dim(col_data), "\n")
cat("Dimensions of counts matrix (Rows x Cols):", dim(counts), "\n")

# N.o samples in the metadata matchs the N.o in the counts
if (nrow(col_data) == ncol(counts)) {
  print("Dimension check passed: Metadata matches the count matrix.")
} else {
  stop("Dimension mismatch: The number of samples in counts does not match the metadata")
}
```

#### **Task 3.10: Construct DESeqDataSet**
```{r}
# Conversion to prevent automatic correction/warning
col_data$condition <- as.factor(col_data$condition)

# Load DESeq2 library
library(DESeq2)

# Construct the DESeqDataSet (dds) object
# countData: The matrix of raw counts
# colData: The dataframe of sample information
# design: The experimental design formula
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = col_data,
                              design = ~ condition)

print(dds)
```

#### **Task 3.11: Transformation**
```{r}
# rlog transformation
rld <- rlog(dds, blind = FALSE)

#  VST transformation
vsd <- vst(dds, blind = FALSE)

# Transformed data in matrix
cat("Dimensions of rlog data:", dim(assay(rld)), "\n")
cat("Dimensions of VST data: ", dim(assay(vsd)), "\n")

# present few rows of transformed values
head(assay(rld), n = 3)
```

#### **Task 3.12: Heatmap of top 40 highly expressed genes**
```{r}
# Load the heatmap library
library(pheatmap)

# Identify  top 40 genes by rlog values
top_genes <- head(order(rowMeans(assay(rld)), decreasing = TRUE), 40)

# Draw heatmap for rlog data
pheatmap(assay(rld)[top_genes, ], 
         main = "Heatmap (rlog): Top 40 Highly Expressed Genes",
         cluster_rows = TRUE, 
         show_rownames = TRUE, 
         cluster_cols = TRUE,
         fontsize_row = 8)

# 3. Draw heatmap for VST data
pheatmap(assay(vsd)[top_genes, ], 
         main = "Heatmap (VST): Top 40 Highly Expressed Genes",
         cluster_rows = TRUE, 
         show_rownames = TRUE, 
         cluster_cols = TRUE,
         fontsize_row = 8)
```

#### **Task 3.13: **
```{r}
library(RColorBrewer)
library(pheatmap)

# Calculate the distance between samples
sampleDists <- dist(t(assay(rld)))

# Convert into a matrix format for plotting
sampleDistMatrix <- as.matrix(sampleDists)

# Define the row and column names (Sample names)
rownames(sampleDistMatrix) <- colnames(rld)
colnames(sampleDistMatrix) <- colnames(rld)

# Set colour palette (Red)
colors <- colorRampPalette(rev(brewer.pal(9, "Reds")))(255)

# 5. Generate heatmap
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors,
         main = "Sample Distance Matrix (SDM) - rlog Data")

```

#### **Task 3.14: Principal Component Analysis using rlog method**
```{r}
library(ggplot2)

# Run PCA and return data for plotting
pcaData <- plotPCA(rld, intgroup = "condition", returnData = TRUE)

# Calculate the percentage of variance for PC1 and PC2
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Print the % significance values
cat("Significance of PC1 (x-axis):", percentVar[1], "%\n")
cat("Significance of PC2 (y-axis):", percentVar[2], "%\n")

# Generate PCA plot using ggplot2
ggplot(pcaData, aes(PC1, PC2, color = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  ggtitle("PCA Plot (rlog method)") +
  theme_minimal()
```

#### **Task 3.15: a.PCA using VST**
```{r}
# 1. Run PCA for VST data
pcaData_vst <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)

# 2. Calculate the percentage of variance for PC1 and PC2
percentVar_vst <- round(100 * attr(pcaData_vst, "percentVar"))

# 3. Print the % significance values for VST
cat("VST - Significance of PC1:", percentVar_vst[1], "%\n")
cat("VST - Significance of PC2:", percentVar_vst[2], "%\n")

# 4. Generate the VST PCA plot
ggplot(pcaData_vst, aes(PC1, PC2, color = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar_vst[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_vst[2], "% variance")) +
  coord_fixed() +
  ggtitle("PCA Plot (VST method)") +
  theme_minimal()
```

#### **Task 3.15: b. Compare PCA using rlog method against VST**



